\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{indentfirst}
\usepackage{microtype}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage[numbers,sort&compress]{natbib}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black,
  pdftitle={What Determines Long-Context Ability in LLMs?},
  pdfauthor={NLP Course Project Draft}
}

\title{\textbf{What Determines Long-Context Ability in LLMs?}\\
\large \textbf{A Survey with Experimental Analysis}}
\author{Sihua Zhang, Minghao Xi}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Long-context ability has become a central theme in modern large language models (LLMs), yet the field still lacks a stable answer to a deceptively simple question: what does it actually mean for a model to \emph{use} long context well? Larger context windows alone are not sufficient. A model may accept 128K or even million-token inputs, while still degrading on retrieval from the middle of a document, on multi-hop aggregation across distant spans, or under realistic serving constraints. This report surveys long-context LLMs through a unified lens that separates five factors: nominal context length, effective retrieval length, long-range reasoning ability, inference-time extension effectiveness, and system serving efficiency. We review architectural routes for long-sequence modeling, context extension methods built on top of Transformer backbones, retrieval and memory augmentation, and system techniques such as IO-aware attention and KV-cache management---including token selection and eviction methods such as H$_2$O, SnapKV, and FIER. We further compare the evaluation landscape, from Needle-in-a-Haystack and position-sensitivity tests to RULER, LongBench, $\infty$Bench, BABILong, L-Eval, and LongBench v2. Finally, we formulate a full experimental plan that is explicitly designed to disentangle whether long-context failure is primarily a retrieval problem, a reasoning problem, an inference-time extension problem, or a serving problem. Our main claim is that long-context ability is not a single scalar property but a joint outcome of model design, prompting and retrieval strategy, task type, and system implementation.
\end{abstract}

\section{Introduction}
Large language models are increasingly deployed in settings that are inherently long-context: document question answering, multi-document synthesis, repository-scale code understanding, long-horizon agent memory, and structured analysis over extended logs or tables. These scenarios have driven rapid progress in nominal context windows, from a few thousand tokens to 32K, 128K, and even million-token regimes in both commercial and open models. However, the field has also accumulated substantial evidence that accepting longer inputs is not equivalent to using them effectively \citep{liu2024lost, hsieh2024ruler, bai2024longbench, kuratov2024babilong, bai2025longbenchv2}.

This mismatch appears in several ways. Position-sensitive evaluations show that models often underuse information placed in the middle of long prompts \citep{liu2024lost}. Synthetic benchmarks beyond single-needle retrieval reveal a significant gap between claimed context length and effective context length \citep{hsieh2024ruler}. Realistic multi-task evaluations suggest that deeper reasoning, rather than nominal window size alone, has become the main bottleneck in long-context settings \citep{bai2025longbenchv2}. At the same time, systems work has shown that long-context deployment is strongly constrained by prefill latency, KV-cache growth, and batching efficiency, even when the underlying model is theoretically capable of longer inputs \citep{dao2022flashattention, kwon2023pagedattention, zheng2024sglang}.

This report is organized around a single question: \emph{what determines long-context ability in LLMs?} Instead of treating long context as one monolithic capability, we decompose it into five dimensions:
\begin{enumerate}[leftmargin=1.5em]
  \item \textbf{Nominal context length}: how many tokens the model and serving stack claim to support.
  \item \textbf{Effective retrieval length}: how far relevant information can be reliably retrieved.
  \item \textbf{Long-range reasoning ability}: whether the model can aggregate, compare, and reason over distant evidence.
  \item \textbf{Inference-time extension effectiveness}: how much performance can be improved without retraining the backbone from scratch.
  \item \textbf{System serving efficiency}: whether the model can be run with acceptable latency, memory, and throughput.
\end{enumerate}

The rest of the report follows this analytical lens. Section~\ref{sec:framework} formalizes the framework. Sections~\ref{sec:modeling}--\ref{sec:systems} survey modeling, extension, retrieval, and systems techniques. Section~\ref{sec:evaluation} reviews benchmarks and what they actually measure. Section~\ref{sec:experiments} proposes a comprehensive experimental design for the course project. We conclude with a discussion of open challenges and a synthesis of current evidence.

\section{Problem Formulation and Analytical Framework}
\label{sec:framework}

\subsection{Why context window size is not enough}
Context window size is best understood as a necessary but insufficient condition. A larger window makes it \emph{possible} to provide more evidence to the model, but it says little about whether the model can retrieve the right facts, combine them correctly, or do so under realistic latency and memory budgets. This distinction has become clearer as long-context evaluations evolved from simple passkey or single-needle retrieval to richer tests of aggregation and multi-hop tracing \citep{hsieh2024ruler, zhang2024infinitebench, kuratov2024babilong}.

\subsection{A five-dimensional view of long-context ability}
We define long-context ability as the model's capacity to preserve useful performance as input length grows along five coupled dimensions.

\paragraph{Nominal context length.}
This is the maximum context size advertised by a model or infrastructure stack. It is influenced by positional encoding, training length, implementation choices, and serving software. It answers ``can the input be admitted?'' but not ``can the input be used well?''

\paragraph{Effective retrieval length.}
This is the largest length regime at which relevant information can still be found reliably. Needle-in-a-Haystack, key-value retrieval, and retrieval-focused RULER tasks are natural measurement tools here \citep{liu2024lost, hsieh2024ruler}.

\paragraph{Long-range reasoning ability.}
This dimension concerns multi-hop chaining, cross-segment aggregation, comparison, and variable tracking over distant evidence. Recent benchmarks increasingly suggest that this is the hardest long-context sub-problem \citep{hsieh2024ruler, kuratov2024babilong, bai2025longbenchv2}.

\paragraph{Inference-time extension effectiveness.}
Many practical approaches do not redesign the architecture from scratch, but instead modify positional scaling, chunk the input, manage cache selectively, or introduce memory-like mechanisms at inference time \citep{chen2023pi, peng2023yarn, ding2024longrope, xiao2023streamingllm, jin2024selfextend, xiao2024infllm, zhang2023h2o, li2024snapkv, wang2025fier}. Their value lies in how much effective context they recover per unit implementation complexity.

\paragraph{System serving efficiency.}
In production and in realistic experiments, long context is constrained by prefill time, decode throughput, peak memory, cache reuse, and batching efficiency. System work is therefore not just a deployment detail; it determines what long-context regimes are empirically reachable \citep{dao2022flashattention, dao2023flashattention2, kwon2023pagedattention, zheng2024sglang}.

\subsection{Main thesis}
The core thesis of this report is that long-context ability is a \emph{joint property} of model architecture, inference strategy, task type, and system realization. A model can be strong on one dimension and weak on another. For example, it may preserve single-needle retrieval but fail on aggregation, or it may reason well at moderate length but become unusably slow under realistic serving constraints.

\section{Technical Foundations of Long-Context Modeling}
\label{sec:modeling}

\subsection{The standard Transformer bottleneck}
The modern discussion starts from the success of the Transformer \citep{vaswani2017attention}. Full self-attention offers flexible content-based interaction and can be written as
\begin{equation}
\mathrm{Attention}(Q,K,V)=\mathrm{softmax}\left(\frac{QK^\top}{\sqrt{d}}\right)V.
\end{equation}
This expression makes the source of the long-context bottleneck explicit: every token can attend to every other token, so the score matrix $QK^\top$ grows with all pairwise interactions. For a sequence of length $n$, this leads to quadratic scaling, which is commonly summarized as
\begin{equation}
\text{time}=\mathcal{O}(n^2 d), \qquad \text{memory}=\mathcal{O}(n^2).
\end{equation}
These asymptotics explain why simply increasing the context window quickly becomes expensive. In long-context regimes, the bottleneck is not only arithmetic complexity; memory traffic and KV-cache growth become equally important \citep{dao2022flashattention}.

\subsection{Sparse and local-global attention}
One classical route is to reduce the interaction graph. Longformer combines local sliding-window attention with task-motivated global tokens, achieving linear scaling in sequence length for many practical settings \citep{beltagy2020longformer}. BigBird augments local attention with random and global connections and shows that sparse structures can preserve strong theoretical properties while enabling much longer inputs \citep{zaheer2020bigbird}. These models demonstrate that many document tasks do not require fully dense attention at every layer and every position.

A simple way to express local attention is to restrict token $i$ to a neighborhood
\begin{equation}
\mathcal{N}(i)=\{j \mid i-w \le j \le i\},
\end{equation}
where $w$ is the window size over past context. The key change is that the attention computation no longer ranges over all $n$ positions, but only over a bounded subset for each token. Under this restriction, the complexity becomes
\begin{equation}
\text{time}=\mathcal{O}(nwd),
\end{equation}
with $w \ll n$ in the long-sequence regime. More general sparse schemes can be viewed as expanding $\mathcal{N}(i)$ with selected global or random connections, but the basic idea is the same: only a limited subset of the full attention graph is retained.

Their main limitation is reduced flexibility in information routing. When the attention pattern is fixed in advance, the model may be unable to directly connect distant tokens that are crucial for a particular instance or task. As a result, sparse and local-global designs primarily improve computational tractability, whereas their effectiveness on complex long-range reasoning depends heavily on how the sparsity pattern is designed and on whether the training setup adequately exposes the model to such dependencies.

\subsection{Linear-time, low-rank, and kernelized alternatives}
Another route is to approximate or reformulate attention itself. This broad family includes low-rank methods, kernelized attention, and recurrent-style sequence models. While these approaches reduce asymptotic cost, the key practical question is whether they retain the content selectivity that makes Transformers powerful in language modeling.

This trade-off is especially clear in recent work on state-space and recurrent models. A standard discrete-time state-space model can be written as
\begin{equation}
h_t = A h_{t-1} + B x_t, \qquad y_t = C h_t,
\end{equation}
where $h_t$ is the latent state, $x_t$ is the input token representation, and $y_t$ is the output. This form highlights the main appeal of state-space models: information is propagated through a compressed recurrent state rather than through an ever-growing pairwise attention map. Mamba argues that earlier linear-time sequence models remained weaker than Transformers on language tasks partly because their sequence dynamics were not sufficiently selective. Its central idea is to make key components of the state-space dynamics input-dependent:
\begin{equation}
(A_t,B_t,C_t)=g(x_t), \qquad
h_t = A_t h_{t-1} + B_t x_t, \qquad y_t = C_t h_t,
\end{equation}
so that the model can modulate what information is retained, updated, or read out at each step \citep{gu2023mamba}. In other words, the recurrence is no longer governed by a fixed transition rule; it is conditioned on the current token. Mamba-2 then reframes the relationship between Transformers and state-space models through structured state space duality, further improving efficiency and implementation simplicity \citep{dao2024mamba2}. Hybrid models such as Jamba suggest that the practical frontier may lie not in replacing attention everywhere, but in combining a small amount of precise attention with more efficient recurrent or state-space computation \citep{lieber2024jamba}. More recent linear recurrent designs such as Gated DeltaNet further strengthen this line by explicitly targeting retrieval and long-context weaknesses in Mamba-like models \citep{yang2024gateddeltanet}.

\subsection{Implication for long-context research}
Architectural work addresses a fundamental question: how can sequence models scale beyond the quadratic regime without giving up too much content-based access? The current evidence suggests a nuanced answer. Pure linear or recurrent models can offer substantial efficiency gains, but the strongest long-context behavior often still benefits from some form of exact or near-exact retrieval path, whether through attention, hybridization, or external memory.

\section{Extending Usable Context Without Retraining the Backbone}
\label{sec:extension}

The previous section focused on architectural changes that modify the sequence model itself. In practice, however, a large share of long-context progress comes from a different strategy: keeping the backbone largely intact while changing how context is represented, accessed, or selected. This family is broader than purely training-free extrapolation. Some methods rely on lightweight fine-tuning or positional recalibration, some reorganize attention during inference, some attach an explicit memory mechanism, and some retrieve only the most relevant evidence before it enters the model context at all. What unifies them is that they aim to expand \emph{usable} context without requiring a completely new backbone.

\subsection{Positional information extension}
Many practical long-context methods keep the Transformer backbone intact and instead modify the positional mechanism. ALiBi demonstrates that a simple linear distance bias can support test-time extrapolation to longer inputs than seen in training \citep{press2022alibi}. At attention head $h$, this can be written as
\begin{equation}
e_{ij}^{(h)}=\frac{q_i^\top k_j}{\sqrt{d}} - m_h(i-j),
\end{equation}
where $m_h$ is a head-specific slope and the bias grows with relative distance. The formula shows that ALiBi does not replace attention itself; instead, it systematically reshapes the attention scores so that distance is encoded directly in the logits. In the RoPE family, Position Interpolation rescales positions back into the original training range, substantially stabilizing long-context extension with minimal fine-tuning \citep{chen2023pi}. A simple way to express this idea is
\begin{equation}
p' = p \cdot \frac{L_{\text{train}}}{L_{\text{test}}},
\end{equation}
where $p$ is a position in the longer test sequence, $L_{\text{train}}$ is the context length seen during training, and $L_{\text{test}}$ is the desired extended length. Intuitively, a long test sequence is compressed back into the positional range where the model is better calibrated. More generally, one can write
\begin{equation}
p' = f(p), \qquad f:[0,L_{\text{test}}] \rightarrow [0,L_{\text{train}}],
\end{equation}
which emphasizes that the extended context is mapped back into the positional range. YaRN improves the efficiency of such extension, while LongRoPE pushes this paradigm to much larger windows through non-uniform interpolation and progressive extension \citep{peng2023yarn, ding2024longrope}.

These methods matter because they are much cheaper than training a new long-context backbone from scratch. However, they mainly answer ``how do we feed longer inputs into the model?'' rather than ``how do we ensure the model reasons well once those inputs are available?'' The distinction is crucial for interpreting benchmark gains.

\subsection{Inference-time long-context adaptation}

\paragraph{Attention-structure reorganization.}
Self-Extend addresses long-context extrapolation by modifying the attention
pattern at inference time. Instead of simply increasing the dense attention
window or discarding distant tokens, it constructs a bi-level attention pattern:
\emph{neighbor attention} preserves fine-grained local interactions within a
short range, while \emph{grouped attention} aggregates farther tokens into
coarser units \citep{jin2024selfextend}. The method therefore tries to preserve
local fidelity while maintaining coarse long-range connectivity, without
requiring further long-context pretraining.

\paragraph{KV-cache quantization, eviction, and retrieval.}
A second line of inference-time methods targets the KV cache budget directly.
These methods do not primarily ask how to change the model's attention structure,
but how to make the stored keys and values cheaper, smaller, or more selective
during generation.

KIVI represents the most direct form of KV-cache compression: it keeps the cache
structure intact but reduces the precision of stored keys and values
\citep{liu2024kivi}. Its central observation is that the key and value caches
have different distributional properties, so they should be quantized
asymmetrically: keys are quantized per channel, while values are quantized per
token. This makes KIVI a useful baseline for KV-cache efficiency because it
compresses the representation of cached information rather than deciding which
tokens to evict.

StreamingLLM instead reduces the active cache by retaining only a small set of
early \emph{attention sink} tokens together with a recent sliding window
\citep{xiao2023streamingllm}. If $\mathcal{S}_{\text{sink}}$ denotes the set of
sink-token indices and $w$ is the recent window size, the retained memory at
time step $t$ can be expressed as
\begin{equation}
\mathcal{M}_t = \mathcal{S}_{\text{sink}} \cup \{t-w+1,\dots,t\}.
\end{equation}
Attention is then computed only over the retained set,
\begin{equation}
\mathrm{Attn}_t=\mathrm{softmax}\!\left(\frac{q_t K_{\mathcal{M}_t}^\top}{\sqrt{d}}\right)V_{\mathcal{M}_t}.
\end{equation}
Compared with KIVI, StreamingLLM changes not the numerical precision of the KV
cache, but the set of tokens that remain visible to the model.

H$_2$O further makes cache retention adaptive. It observes that a small subset
of tokens, termed \emph{heavy hitters}, accumulate disproportionately large
attention scores and are repeatedly referenced during decoding
\citep{zhang2023h2o}. The cache therefore balances heavy-hitter tokens with a
recent window. The eviction decision can be framed as selecting a subset
$\mathcal{C}_t \subset \{1,\dots,t\}$ of size $B$ that maximizes accumulated
attention mass:
\begin{equation}
\mathcal{C}_t = \operatorname*{arg\,max}_{|\mathcal{S}|=B} \sum_{i \in \mathcal{S}} \sum_{s \le t} a_{s,i},
\end{equation}
where $a_{s,i}$ is the attention score from decoding step $s$ to token $i$.
This formulation makes the selection criterion explicit: tokens that have been
consistently attended to are preferred over tokens that were only briefly
relevant.

SnapKV shifts the selection decision to the prefill phase rather than performing
it incrementally during decoding \citep{li2024snapkv}. The key observation is
that each attention head exhibits a predictable focus pattern: the positions it
will attend to during generation can be inferred from a short \emph{observation
window} at the end of the prompt. SnapKV uses this window to score and cluster
KV positions per head, retaining only high-importance positions before
generation begins. This front-loaded compression avoids the overhead of
per-step eviction.

FIER pushes the granularity of selection further \citep{wang2025fier}. Rather
than operating at the page or chunk level, it selects individual tokens from the
full cache at each decoding step. To make token-level retrieval efficient, FIER
uses 1-bit quantized keys as lightweight importance proxies, enabling
query-relevant token selection without the cost of full-precision key
comparisons.

Taken together, these methods form a spectrum of KV-cache adaptation strategies.
KIVI compresses the representation of the cache; StreamingLLM applies a fixed
sink-plus-window retention rule; H$_2$O performs dynamic heavy-hitter eviction;
SnapKV front-loads selection into the prefill stage; and FIER performs
fine-grained query-aware token retrieval. They differ in \emph{what} is reduced
(precision or token set), \emph{when} selection occurs, and \emph{how}
importance is estimated.

\subsection{Retrieval as context selection for long-context processing}

So far, the discussion has focused on inference-time methods that reorganize
attention or reduce the active KV cache. A complementary view is that
long-context processing is often less about making every token visible at all
times, and more about deciding which pieces of context should be surfaced for
the current computation. Under this view, retrieval is a form of
\emph{context selection}: it constructs a more useful working context before, or
during, model reasoning.

InfLLM sits at the boundary between KV-cache adaptation and retrieval-based
long-context processing. It stores distant context into additional memory units
and retrieves token-relevant units during attention computation
\citep{xiao2024infllm}. Unlike the eviction methods above, which decide which
cached keys and values should remain active within the model's standard attention
mechanism, InfLLM externalizes remote context into a structured memory and
selectively brings back relevant units when needed. The distinction matters: KV
eviction methods operate entirely within the attention computation and are
transparent to the rest of the model, whereas InfLLM introduces an explicit
external memory that the model queries. This makes InfLLM closer in spirit to
retrieval-augmented approaches, even though it operates at inference time without
any offline indexing step.

That boundary---between inference-time memory management and offline retrieval
augmentation---is worth making explicit. The methods discussed so far (KIVI,
StreamingLLM, H$_2$O, SnapKV, FIER, InfLLM) all operate at generation time on
a single input: they decide how to compress, evict, or selectively access the
KV cache for the current sequence. Retrieval-augmented generation, by contrast,
typically involves a separate retrieval step that runs before the model sees the
query, drawing from an indexed corpus that is independent of the current
generation context. This is a fundamentally different intervention point in the
long-context pipeline.

This perspective is especially important in multi-document QA, enterprise
knowledge access, codebase understanding, and long-form scientific synthesis,
where reading everything at once may be either too expensive or too distracting.
LongBench already showed that retrieval and compression can help weaker
long-context models, although they do not fully substitute for intrinsically
stronger long-context understanding \citep{bai2024longbench}. More recent work
rethinks the granularity and structure of retrieval itself. RAPTOR organizes
documents through recursive abstraction for tree-structured retrieval
\citep{sarthi2024raptor}. LongRAG argues that stronger readers make much larger
retrieval units feasible and sometimes preferable \citep{wang2024longrag}.
Anthropic's contextual retrieval further emphasizes that retrieval quality
depends on preserving local context around chunks, rather than embedding
isolated fragments alone \citep{anthropic2024contextualretrieval}.

The key point for this section is that long-context capability is not always
achieved by passing more tokens through the model in a single shot. In many
settings, it is achieved by deciding which evidence should be surfaced, in what
granularity, and in what organization, before the model reasons over it.
\subsection{What these strategies actually change}
The methods surveyed in this section should not be grouped together merely because they are all cheaper than retraining a new backbone. They intervene at different points in the long-context pipeline and address different failure modes.

Positional extension methods (ALiBi, PI, YaRN, LongRoPE) change \emph{admissible length}: they determine how long an input the model can accept without positional out-of-distribution failure, but say nothing about whether the model uses that input well.

Attention-structure reorganization (Self-Extend) changes \emph{how the model attends} to distant tokens during inference, trading off local precision against coarse long-range coverage.

KV-cache adaptation methods (KIVI, StreamingLLM, H$_2$O, SnapKV, FIER) change \emph{what is retained in memory} during generation. They differ in whether they compress representation precision (KIVI), apply a fixed retention rule (StreamingLLM), perform dynamic importance-based eviction (H$_2$O), front-load selection into prefill (SnapKV), or perform fine-grained per-step token retrieval (FIER).

Inference-time memory augmentation (InfLLM) changes \emph{how remote context is stored and accessed}, externalizing it beyond the standard KV cache while remaining tied to the current generation.

Retrieval augmentation (RAPTOR, LongRAG, contextual retrieval) changes \emph{which evidence enters the context at all}, operating on an offline-indexed corpus before generation begins.

Because these strategies target different failure modes, they should not be evaluated with a single metric, and gains on one dimension do not imply gains on others.

\section{Systems for Long-Context LLMs}
\label{sec:systems}

Long-context research often foregrounds architecture and benchmark accuracy, but real deployment is frequently dominated by system constraints. If prefill is too slow, if KV-cache memory prevents useful batching, or if cache reuse is poor across repeated prefixes, then nominal context support has limited practical value.

\subsection{IO-aware attention and kernel optimization}
FlashAttention reframed attention as an IO problem, showing that reducing high-bandwidth-memory traffic yields large practical gains without approximating the attention computation \citep{dao2022flashattention}. FlashAttention-2 further improves parallelism and work partitioning, pushing exact attention closer to hardware limits \citep{dao2023flashattention2}. These results matter for long context because they lower the constant factors that otherwise make even exact attention impractical at scale.

\subsection{KV-cache management and serving engines}
PagedAttention and vLLM address a central bottleneck in autoregressive serving: the dynamic and memory-hungry KV cache. For a Transformer with $N_{\text{layers}}$ layers, $H$ heads, head dimension $d_{\text{head}}$, sequence length $L$, and per-element storage cost $b$ bytes, the KV-cache size can be approximated as
\begin{equation}
M_{\text{KV}} \approx 2 \cdot L \cdot N_{\text{layers}} \cdot H \cdot d_{\text{head}} \cdot b,
\end{equation}
where the factor of $2$ accounts for keys and values. This estimate clarifies why long-context serving rapidly becomes memory-bound even when model weights are fixed: the cache grows linearly with sequence length and with the depth of the model. A complementary high-level view is that, for standard Transformers, prefill scales roughly as
\begin{equation}
\text{Prefill cost} \sim \mathcal{O}(L^2), \qquad
\text{Decode step cost} \sim \mathcal{O}(L),
\end{equation}
which explains why long prompts and long generation histories stress the system in different ways: prefill is dominated by prompt length, while decoding remains sensitive to the accumulated context. By managing KV memory like virtual memory pages, vLLM substantially improves throughput, especially at longer sequence lengths \citep{kwon2023pagedattention}. SGLang extends the systems perspective beyond one-shot decoding, introducing RadixAttention for efficient prefix reuse across structured language-model programs \citep{zheng2024sglang}. This is especially relevant for agentic and multi-step pipelines, where repeated long prefixes are common.

\subsection{Serving Efficiency as a Dimension of Long-Context Ability}

The system perspective is not only about how to deploy a long-context model more efficiently; it also changes what should count as long-context ability in the first place. In the analytical framework of this report, serving efficiency is one of the core dimensions of long-context performance rather than a secondary implementation detail. The relevant metrics include time-to-first-token, prefill latency, decode throughput, peak memory, batch scalability, and cache reuse efficiency. These quantities determine which context lengths remain usable in practice, under what cost, and for what type of workload. A model that performs well on long-context benchmarks but becomes prohibitively slow or memory-intensive at deployment time does not fully realize its nominal long-context capability. For this reason, system efficiency should be evaluated alongside retrieval and reasoning rather than after them.


\section{Evaluation of Long-Context Ability}
\label{sec:evaluation}

\subsection{From retrieval sanity checks to deeper reasoning tests}
Evaluation has matured from simple retrieval tests to increasingly realistic multi-task reasoning benchmarks. Needle-in-a-Haystack and related key-value retrieval tasks are useful sanity checks, but they are too narrow to define long-context ability by themselves \citep{liu2024lost}. Indeed, position-sensitivity analysis shows that even models with long context windows may systematically underuse information placed in the middle of prompts \citep{liu2024lost}.

\subsection{Synthetic but harder benchmarks}
RULER is a major step beyond single-needle testing. It introduces configurable synthetic tasks spanning retrieval, aggregation, and tracing, and explicitly emphasizes the distinction between claimed and effective context size \citep{hsieh2024ruler}. $\infty$Bench pushes evaluation beyond 100K tokens and combines synthetic and realistic tasks, highlighting that current models still struggle in truly long regimes \citep{zhang2024infinitebench}. BABILong makes the reasoning challenge sharper still by scattering relevant facts through extremely long natural text and measuring whether models can chain them reliably \citep{kuratov2024babilong}.

\subsection{Realistic long-context benchmarks}
LongBench provides a bilingual, multi-task benchmark spanning question answering, summarization, few-shot learning, synthetic tasks, and code completion \citep{bai2024longbench}. L-Eval contributes not only datasets but also a methodological warning: traditional n-gram overlap is often misaligned with long-form answer quality, so long-context evaluation may require stronger judges and better instructions \citep{an2023leval}. LongBench v2 raises the bar further by focusing on realistic multi-task problems that demand deeper understanding and reasoning, suggesting that the next bottleneck is not merely feeding longer documents, but reasoning effectively once they are available \citep{bai2025longbenchv2}.

\subsection{Benchmark-to-capability mapping}
An important methodological lesson is that benchmarks measure different slices of long-context ability:
\begin{itemize}[leftmargin=1.5em]
  \item \textbf{Needle-in-a-Haystack / key-value retrieval}: basic retrieval and position robustness.
  \item \textbf{Lost in the Middle}: positional sensitivity under long inputs.
  \item \textbf{RULER retrieval tasks}: effective retrieval length.
  \item \textbf{RULER aggregation and tracing}: long-range reasoning over synthetic evidence.
  \item \textbf{LongBench / LongBench v2}: realistic long-context understanding and reasoning.
  \item \textbf{Serving benchmarks}: practical utility under latency and memory constraints.
\end{itemize}

This mapping is critical for experiment design. Without it, one may overgeneralize from a benchmark that measures only one sub-skill. It is therefore useful to make the notion of effective context explicit. If $\mathrm{Acc}(L)$ denotes performance at context length $L$, then one natural definition is
\begin{equation}
L_{\mathrm{eff}}(\tau)=\max \{L:\ \mathrm{Acc}(L)\ge \tau\},
\end{equation}
where $\tau$ is a task-dependent target threshold such as $0.8$. This definition captures the longest context regime at which performance remains above a chosen quality bar, and therefore formalizes the difference between claimed window size and usable window size. To quantify how sharply performance degrades with length, one may also track a decay slope such as
\begin{equation}
\Delta(L_1,L_2)=\frac{\mathrm{Acc}(L_2)-\mathrm{Acc}(L_1)}{\log L_2-\log L_1}.
\end{equation}
This second quantity is useful because two models may have similar accuracy at one length while differing substantially in how rapidly they deteriorate as the context grows.

\section{Experimental Analysis Plan}
\label{sec:experiments}

\subsection{Goal of the experimental section}
The experimental section of the course project should not aim to produce another generic leaderboard. Its purpose is to answer a sharper question: \emph{when long-context performance fails, which factor fails first?} To make that question operational, we propose a complete experiment matrix that disentangles retrieval, reasoning, inference-time extension, and system serving.

\subsection{Research questions}
We recommend structuring the experiments around the following research questions:
\begin{description}
  \item[RQ1] Does long-context degradation appear first as retrieval failure or as reasoning failure?
  \item[RQ2] How much can inference-time extension improve effective context length without retraining the base model?
  \item[RQ3] Do different benchmarks agree on what it means for a model to be good at long context?
  \item[RQ4] Under realistic serving setups, do latency, memory, and throughput become the dominant bottlenecks before benchmark accuracy does?
\end{description}

\subsection{Recommended model and system choices}
A practical primary model is a 7B--8B instruction-tuned checkpoint with an established ecosystem, such as Llama 3.1 8B Instruct or a Qwen2.5 7B long-context variant. For system comparisons, the most informative trio is:
\begin{enumerate}[leftmargin=1.5em]
  \item a vanilla Hugging Face Transformers baseline,
  \item vLLM for PagedAttention-based serving,
  \item SGLang for structured execution and prefix reuse.
\end{enumerate}
This combination is strong because it keeps the model fixed while changing the serving substrate.

\subsection{Experiment family I: retrieval-focused evaluations}
The first family should cover basic retrieval and its failure patterns:
\begin{itemize}[leftmargin=1.5em]
  \item Needle-in-a-Haystack across context lengths such as 4K, 8K, 16K, 32K, 64K, and 128K.
  \item Needle placement at multiple positions, including beginning, middle, and end.
  \item RULER retrieval or multi-key retrieval tasks for a harder synthetic setting.
\end{itemize}
Expected outcome: basic retrieval remains relatively strong longer than reasoning, but degrades with unfavorable position and increasing length.

\subsection{Experiment family II: reasoning-focused evaluations}
The second family should explicitly separate retrieval from reasoning:
\begin{itemize}[leftmargin=1.5em]
  \item RULER aggregation tasks.
  \item RULER tracing or variable tracking tasks.
  \item A small realistic subset such as LongBench QA or selected multi-document QA datasets.
\end{itemize}
Expected outcome: performance should degrade earlier and more steeply than on basic retrieval, even when the relevant information is technically present in context.

\subsection{Experiment family III: inference-time extension}
The third family examines methods that do not require retraining the full backbone:
\begin{itemize}[leftmargin=1.5em]
  \item baseline direct inference,
  \item one positional extension approach, such as a RoPE scaling or interpolation variant,
  \item one training-free memory or extrapolation method, such as Self-Extend or InfLLM,
  \item one KV cache token selection method, such as SnapKV or FIER,
  \item optionally a streaming-oriented method such as StreamingLLM.
\end{itemize}
The key dependent variables should include effective context length, decay slope across lengths, and whether gains appear mostly on retrieval or also on reasoning. KV cache selection methods are particularly informative here because they expose a different trade-off from positional extension: rather than asking ``how far can the model reach?'', they ask ``given a fixed memory budget, which tokens should be kept?'' Comparing SnapKV (prefill-time selection) against FIER (per-step token-level retrieval) can further reveal whether the timing and granularity of selection matter for task accuracy. For task accuracy itself, the simplest reference metric is
\begin{equation}
\mathrm{Acc}=\frac{1}{N}\sum_{i=1}^{N}\mathbf{1}\{\hat{y}_i=y_i\},
\end{equation}
which is appropriate when the answer space is discrete or when exact match is the desired success criterion. In long-context settings, task-specific alternatives such as normalized match or substring match are often also needed, especially when answers are free-form but semantically simple.

\subsection{Experiment family IV: system serving}
The fourth family focuses on the serving stack itself:
\begin{itemize}[leftmargin=1.5em]
  \item time-to-first-token,
  \item prefill latency,
  \item decode throughput,
  \item peak memory,
  \item batch scalability,
  \item repeated-prefix speedup when prefix caching is available.
\end{itemize}
The experimental comparison should keep model, prompt, and decoding settings fixed while varying only the serving engine. This isolates the systems contribution cleanly.

\subsection{Suggested outputs}
The report should ultimately present:
\begin{itemize}[leftmargin=1.5em]
  \item heatmaps for retrieval accuracy by length and position,
  \item curves of accuracy versus length for retrieval and reasoning tasks,
  \item effective context length comparisons across inference-time methods,
  \item throughput and latency curves for different serving engines,
  \item a unified analysis table summarizing where each failure mode becomes dominant.
\end{itemize}

\section{Discussion and Open Challenges}

\section{Conclusion}


\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
